Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-01 anzjs2018/sections/03_models.tex
edits to sec 3

The particle filter allows us to model several important bus behaviours:
\begin{itemize}
\item non-constant speed along roads,
\item optional stopping and waiting at bus stops while passengers board and disembark, and
\item optional stopping and waiting at intersections or traffic lights (future work).
\end{itemize}

For each particle, the transition function generates a plausible trajectory,
using system noise parameter $\sigma^2$ which describes 
how the vehicle's speed changes as a Markov process.
The vehicle does not travel constantly along the route,
however, as it needs to service bus stops along the way,
so the transition function includes 
stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
board and disembark (conditional on the bus stopping),
and the minimum dwell time at stops, $\gamma$,
for the doors to open and close,
and account for deceleration and acceleration.
For these parameters, we used constant values for all stops,
based on those used by \cite{Hans_2015};
future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
which implements those features discussed above.
The stop positions are represented by their distance into trip,
$\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
the $stop\_index$ function returns the index of a stop,
given a distance travelled and set of stop distances $\boldsymbol{S}$.
Lines~2--9 are necessary to decrease the degeneration rate of the filter,
which can occur when there is a long delay between observations
simply because the bus has not moved due to traffic or other reasons. 
We therefore place positive probability for some particles to remain stationary.

Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
which involves changing the vehicle's speed each second (i.e., acceleration).
Lines~13--25 handle arrival behaviour at stops:
if the particle passes the next stop,
it has a chance of stopping (lines~18--19),
in which case the dwell time is sampled and subtracted from the 
time remaining to travel (lines~20--21).
Finally, to track travel time along segments,
at each point where the vehicle travels (lines 7 and 26),
the travel time for the current segment, $z_j$, is incremented.
The mutation ends once the travel time $\delta$ reaches zero,
or when the vehicle reaches the end of the route (lines 10, 14).
Once complete, the particle's new state is returned (line 29).

@@

After mutating the particle set, the posterior distribution $\tilde\bX_{k|k}$ 
is obtained by reweighting each particle using the likelihood.
The measurement function $h$,
and an additional function $g$ which transforms GPS coordinates onto a flat
surface (the geographical equirectangular projection, \cite{Snyder_1998}),
allow the placement of a bivariate normal likelihood on the data.
The model for the observations is,
assuming GPS error $\epsilon^2$,
\begin{equation*}
\label{eq:pf_obs_model}
g(\bY_k) = g(h(\bX_k)) + \br_k,
\quad \br_k \sim \mathrm{N}(\boldsymbol{0}, \epsilon^2\boldsymbol{I})
\end{equation*}
The geographical distance between the observation and the true vehicle can be expressed
in terms of two independent random variables $z_1, z_2 \sim \mathrm{N}(0,1)$
\begin{equation*}
\label{eq:obs_dist}
dist(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\epsilon z_1)^2 + (\epsilon z_2)^2}.
\end{equation*}
The sum of two independent, standard normal random variables 
is $\chi^2$ distributed with 2~degrees of freedom,
which is itself exponential,
so rearranging gives
\begin{equation*}
\label{eq:obs_exp}
\left(\frac{dist(\bY_k, h(\bX_k))}{\epsilon}\right)^2 =
z_1^2 + z_2^2 \sim \mathrm{Exp}\left(\frac{1}{2}\right).
\end{equation*}

The likelihood of the data given a particle's state estimate 
can now be calculated as
\begin{equation*}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2}\exp\left\{
-\frac{1}{2} \left(\frac{dist(\bY_k, h(\bX_k^{(i)}))}{\epsilon}\right)^2
\right\}
\end{equation*}
It is worth noting that this representation of the likelihood is only
possible due to the discrete nature of the particle filter state estimate;
in other approaches, such as the Kalman filter,
a reverse non-deterministic transformation,
the inverse measurement function,
is required to estimate the ``observed'' distance travelled $\hat Z_k~=~h^{-1}(\bY_k)$, 
which introduces additional error and uncertainty into the model.

>>>>

The particle filter allows us to flexibly model bus behaviour,
which in our current model includes
\begin{itemize}
\item non-constant speed along roads (between observations), and
\item bus stop behaviour, which involves optional stopping and waiting times
    while passengers board and disembark.
\end{itemize}
To handle the above, the transition function consists of two models,
one for each of the above behaviours.
The first models the speed and distance travelled of the vehicle 
by using Newton's Laws of Motion,
which, letting $\delta = t_k - t_{k-1}$ and $\epsilon_k^{(i)}\sim\mathrm{N}(0, \sigma^2)$, is
\begin{equation}
\label{eq:trans_newton}
\bX_k^{(i)} = f_1(\bX_{k-1}^{(i)}, \delta_k, \epsilon_k^{(i)}) = 
    \begin{bmatrix}
        x_{k-1}^{(i)} + \delta(\dot x_{k-1}^{(i)} + \epsilon_k^{(i)}) \\
        \dot x_{k-1}^{(i)} + \epsilon_k^{(i)}
    \end{bmatrix}.
\end{equation}

The second vehicle behaviour requires a model of bus stop wait times.
We assume the vehicle is approaching stops $j$,
and will stop with some probability $\pi_j$.
If it does not stop, the remaining travel time $\delta_k$ is unaltered;
otherwise, the bus waits for a minimum of $\gamma$ seconds---this 
accounts for deceleration, opening and closing of doors, and acceleration
\citep{Hands_2015}---and then waits while passengers board and disembark,
which is assumed to be exponentially distributed with mean $\tau_j$.
This leads to the following model,
\begin{equation}
\label{eq:trans_stop}
\begin{split}
p_j^{(i)} &\sim \mathrm{Bernoulli}(\pi_j) \\
\tilde d_j^{(i)} &\sim \mathrm{Exponential}(\tau_j) \\
d_j^{(i)} &= p_j^{(i)}(\gamma + \tilde d_j^{(i)}).
\end{split}
\end{equation}

The combination of (\ref{eq:trans_newton}) and (\ref{eq:trans_stop})
is handled by the transition function algorithm,
displayed in Algorithm~\ref{fig:algorithm}.
While (\ref{eq:trans_stop}) is used directly in the algorithm,
(\ref{eq:trans_newton}) is modified to allow the bus to approach a stop
and conditionally wait before continuing,
until $\delta_k$ reaches zero.
The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
along each segment $\ell$ of the route,
storing the times so that the posterior distribution of the travel time
can be computed once all particles have completed segment $\ell$:
\begin{equation}
\label{eq:post_tt}
p(z_\ell | \bY_{1:k}) \approx
    \sum_{i=1}^N W_k^{(i)} \dot\delta_{z_\ell}(z_\ell^{(i)}).
\end{equation}

% For each particle, the transition function generates a plausible trajectory,
% using system noise parameter $\sigma^2$ which describes 
% how the vehicle's speed changes as a Markov process.
% The vehicle does not travel constantly along the route,
% however, as it needs to service bus stops along the way,
% so the transition function includes 
% stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
% dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
% board and disembark (conditional on the bus stopping),
% and the minimum dwell time at stops, $\gamma$,
% for the doors to open and close,
% and account for deceleration and acceleration.
% For these parameters, we used constant values for all stops,
% based on those used by \cite{Hans_2015};
% future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

% The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
% which implements those features discussed above.
% The stop positions are represented by their distance into trip,
% $\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
% the $stop\_index$ function returns the index of a stop,
% given a distance travelled and set of stop distances $\boldsymbol{S}$.
% Lines~2--9 are necessary to decrease the degeneration rate of the filter,
% which can occur when there is a long delay between observations
% simply because the bus has not moved due to traffic or other reasons. 
% We therefore place positive probability for some particles to remain stationary.

% Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
% which involves changing the vehicle's speed each second (i.e., acceleration).
% Lines~13--25 handle arrival behaviour at stops:
% if the particle passes the next stop,
% it has a chance of stopping (lines~18--19),
% in which case the dwell time is sampled and subtracted from the 
% time remaining to travel (lines~20--21).
% Finally, to track travel time along segments,
% at each point where the vehicle travels (lines 7 and 26),
% the travel time for the current segment, $z_j$, is incremented.
% The mutation ends once the travel time $\delta$ reaches zero,
% or when the vehicle reaches the end of the route (lines 10, 14).
% Once complete, the particle's new state is returned (line 29).

@@

To compute the likelihood of the observation given each particle state,
$p(\bY_k|\bX_k^{(i)})$,
the measurement function $h$ is needed along with an
\emph{geographical projection} $g$ to allow comparison of $\bY_k$,
a GPS observation, with $\bX_k$, a distance travelled along the route in meters.

The measurement function $h$ computes the GPS coordinates by using the 
shape information provided by GTFS and travelling $x_k$ meters along 
this two-dimensional line.
Meanwhile, we use the \emph{Equirectangular projection} \citep{Snyder_1998},
$\br = g(\bY_1 | \bY_0)$,
so that the magnitude of $\boldsymbol{r}$ is equal to the physical distance
between $\bY_0$ and $\bY_1$, with both dimensions in meters.
Let $\bY_i = [\lambda_i, \psi_i]^\top$,
with $\lambda_i$ and $\psi_i$ measured in radians,  
and $R$ is the Earth's radius, in meters, then
\begin{equation}
\br = 
g(\bY_1 | \bY_0) = 
    \begin{bmatrix}
        r_1 \\ r_2
    \end{bmatrix} =
    R \begin{bmatrix}
        (\lambda_1 - \lambda_0)\cos\phi_0 \\
        \phi_1 - \phi_0
    \end{bmatrix}
\end{equation}
and, more importantly, the geographical distance $\mathcal{D}$ between the two 
GPS coordinates is given by
\begin{equation}
\label{eq:dist}
\mathcal{D}(\bY_0, \bY_1) = ||g(\bY_0 | \bY_1)|| = \sqrt{r_1^2 + r_2^2}
\end{equation}

Now we assume GPS observations have a known error of $\omega^2$ meters,
and that \mbox{$\br \sim \mathrm{N}(\boldsymbol{0}, \omega^2\mathbf{I})$},
which allows us to express $\br$ in terms of two independent
standard normal random variables $z_1, z_2 \sim \mathrm{N}(0,1)$,
which lets us express the distance between a vehicle observation $\bY_k$
and a state $\bX_k$ using (\ref{eq:dist}) and the measurement function $h$,
\begin{equation}
\label{eq:obs_dist}
\mathcal{D}(\bY_k, h(\bX_k)) = \sqrt{r_{k1}^2 + r_{k2}^2} 
    = \sqrt{(\omega z_1)^2 + (\omega z_2)^2}
    = \sqrt{\omega^2 (z_1^2 + z_2^2)}
\end{equation}

Finally, it is known that the distribution of the sum of two indendent 
standard normal variables is exponentially distributed with rate $\frac{1}{2}$,
and that if $Z \sim \mathrm{Exponential}(\theta)$ then
$cZ \sim \mathrm{Exponential}(\frac{\theta}{c})$,
so the \emph{squared} distance from (\ref{eq:obs_dist}) has exponential distribution
\begin{equation}
\label{eq:obs_exp}
\mathcal{D}(\bY_k, h(\bX_k))^2 =
\omega^2(z_1^2 + z_2^2) \sim \mathrm{Exponential}\left(\frac{1}{2\omega^2}\right).
\end{equation}

The likelihood of the observation $\bY_k$ given a state estimate $\bX_k^{(i)}$
can now be expressed using (\ref{eq:obs_exp}) and the PDF of the exponential distribution,
\begin{equation}
\label{eq:lhood}
p(\bY_k | \bX_k^{(i)}, \epsilon) =
\frac{1}{2\omega^2}\exp\left\{
    -\frac{\mathcal{D}(\bY_k, h(\bX_k^{(i)}))^2}{2\omega^2}
\right\},
\end{equation}
where the distance between two GPS coordinates is easily computed,
making (\ref{eq:lhood}) easy and fast to evaluate within the particle filter algorithm.

It is worth noting that this representation of the likelihood is only
possible due to the discrete nature of the particle filter state estimate;
in the Kalman filter, which has often been used in vehicle tracking,
the measurement \emph{matrix} is used, and required that there is a linear
transformation between the state and its observations.
To enable this, applications first estimate the \emph{observed distance travelled}
by snapping the GPS observations to the route,
which introduces unnecessary error and uncertainty into the model.
Our approach avoids this, which makes it more stable in locations where two 
parts of the route are close to each other,
such as at loops, where a single GPS observation might have two likely ``snapping'' points.
